SampleSizePlanner: A Tool to Estimate and Justify Sample Size for Two-Group Studies

### 2021-07-16 inst/manuscript/manuscript.Rmd
updated manuscript according to the first round of revision

---
title             : "SampleSizePlanner: A Tool to Estimate and Justify Sample Size for Two-Group Studies"
shorttitle        : "SAMPLESIZEPLANNER: A TOOL TO ESTIMATE AND JUSTIFY SAMPLE SIZE"

@@

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
linkcolor         : "blue"
mask              : no
draft             : no
figsintext        : yes
numbersections    : true

params:
  highlight_changes: TRUE

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_docx
---

@@

# Introduction {.unnumbered}

@@

In the statistical literature, a few methods have been proposed to determine and justify sample size. In practice, however, authors are short of practical guides on how to navigate among the different sample size methods. The aim of our tutorial is to point out for each method the essential decision points that a researcher has to face during this process. We also provide a collection of ready-to-use analysis code and a ShinyApp that helps researchers use and report the main sample size estimation techniques for different scenarios. The tutorial is focused exclusively on the scenario of the comparison of two independent groups (i.e., the independent *t*-test design) with a one-sided test.

# Sample Size Determination and Justification {.unnumbered}

@@

## Choosing a method in case of sample size justification {.unnumbered}

@@

Within the realm of statistical testing, there are some other factors that affect the preferred inference tool: Do you prefer to test for equivalence (no difference in mean) or for superiority (mean of one group larger than mean of other group), are you interested in calculating a required sample size for a specific hypothetical effect size or for a range of possible values, and do you wish to employ sequential testing (applicable to Bayesian testing)? For frequentist estimation, the preferred inference tool might differ depending on whether we evaluate uncertainty for each group separately or jointly. We will describe these specific factors when we go into detail about each of the preferred methods. A flow-chart representing all of these choices is given in Figure \@ref(fig:flowchart).

@@

## How to use this guide {.unnumbered}

@@

For each method, only the main parameters can be adjusted with a certain range of values in the ShinyApp by using a slider. These parameters are presented in the text in bold. Other parameters are set to preset values in the application but can be adjusted in the accompanying R package to any sensible value. These parameters are highlighted in italics in the tutorial. Both the app and the package allow the users to save or copy a text template with the results of the sample size determination. We offer a list of possible justifications at the decision points for each method (indicated between square brackets), but users are able to provide their own justification as free-text. The provided justification text could serve as a stub for the description of the chosen sample size in a paper, a preregistration or registered report, or a grant proposal.

Throughout, we will use the example story of Mary the educational psychologist. Mary has come up with a new set of games that challenge spatial insight. She believes that distributed and targeted engagement with these games for a period of six months for children in the age range of 8 to 12 will lead to lasting improvements on their IQ score as measured through Raven's progressive matrices test (population mean 100, population SD 15). Mary collects data for a control sample that gets regular education and for an experimental sample and plans to compare those samples. For illustrative purposes, we will have Mary's hypothetical study goal depend on the earlier presented decisions to illustrate different scenarios for each sample size planning method.

@@

## 1. Testing {.unnumbered}

### 1.1. Effect size = 0 {.unnumbered}

#### 1.1.1. Two One‐Sided Tests (TOST) {.unnumbered}

##### Study context {.unnumbered}

@@

##### Description {.unnumbered}

TOST is a frequentist equivalence testing approach that adopts two one-sided hypotheses to designate an interval hypothesis [@schuirmann_comparison_1987]. The lower and upper boundaries of the interval are determined by the equivalence band (i.e. SESOI) around the expected population effect size (e.g., 0). @lakens_equivalence_2018 lists several methods that can be used to determine the SESOI . In case of TOST, the two null hypotheses state that the effect size is equal to the lower and upper equivalence band values, whereas the alternative hypotheses state that the effect size is significantly smaller than the upper equivalence band value and significantly larger than the lower equivalence band value. In case both one-sided tests reject the null-hypotheses at a given significance level, the group means are considered to be practically equivalent. See @lakens_equivalence_2018, for further reading.

##### Parameters {.unnumbered}

@@

##### How to use the package {.unnumbered}

@@

##### How to report your sample size estimation {.unnumbered}

In order to calculate an appropriate sample size for testing whether the two groups are practically equivalent, we used the Two One-Sided Tests of Equivalence [TOST; @schuirmann_comparison_1987] method. We set the aimed TPR to be 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. We consider all effect sizes below 0.2 equivalent to zero, because [1) previous studies reported a similar equivalence band; 2) of the following substantive reasons: ...]. The expected delta was 0 because [1) we expected no difference between the groups]. Based on these parameters, a sample size of 429 per group was estimated in order to reach a TPR of 0.8 with our design.

#### 1.1.2. Equivalence interval Bayes factor {.unnumbered}

##### Study context {.unnumbered}

@@

##### Description {.unnumbered}

Equivalence interval Bayes factors contrast an equivalence hypothesis to a non-equivalence hypothesis and quantify the evidence with Bayes factors. Typically, H~0~ constitutes the equivalence interval (comparable to SESOI in the TOST framework), and H~a~ constitutes the complementary non-equivalence regions. Formally, the Bayes factor is calculated by dividing the fraction *posterior area inside the interval/posterior area outside the interval* (i.e., the posterior odds) by the fraction *prior area inside the interval/prior area outside the interval* (i.e., the prior odds). The resulting value quantifies how much more likely it is that the data occurred under a population effect size deemed 'equivalent' relative to the data having occurred under a population effect size deemed non-equivalent. The current implementation uses a default Cauchy prior on effect size with scale parameter 1/$\sqrt{2}$. For further reading, see @morey_bayes_2011 and @van_ravenzwaaij_bayes_2019.

##### Parameters {.unnumbered}

**Delta:** The expected population effect size.\
**TPR:** The desired long-run probability of obtaining a Bayes factor at least as high as the Threshold, given Delta.\
**EqBand:** The chosen width of the equivalence region.\
*Threshold:* Critical threshold for the Bayes factor. The threshold level in the application can be set to 10, 6, or 3.\

##### How to use the package {.unnumbered}

```{r, eval = FALSE, tidy = TRUE, echo = TRUE}
SampleSizePlanner::ssp_eq_bf(
  tpr = 0.8, delta = 0,
  eq_band = 0.2, thresh = 10
  )
```

##### How to report your sample size estimation {.unnumbered}

In order to estimate the sample size, we used the interval equivalent Bayes factor [@morey_bayes_2011; @van_ravenzwaaij_bayes_2019] method. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. We consider all effect sizes below 0.2 equivalent to zero, because [1) previous studies reported a similar equivalence region; 2) of the following substantive reasons: ...]. The expected delta was 0 because [1) we expected no difference between the groups]. Our Bayes factor threshold for concluding equivalence was 10. Based on these parameters, a minimal sample size of 144 per group was estimated in order to reach 0.8 TPR for our design.

### 1.2. Effect size \>0 {.unnumbered}

#### 1.2.1. Frequentist {.unnumbered}

##### 1.2.1.1. Classical power analysis {.unnumbered}

###### Study context {.unnumbered}

@@

###### Description {.unnumbered}

The classical power analysis approach allows to calculate the required sample size in order to obtain a significant result for the null hypothesis test a certain proportion of times in the long run given an assumed population effect size.

###### Parameters {.unnumbered}

@@

###### How to use the package {.unnumbered}

@@

###### How to report your sample size estimation {.unnumbered}

We used a power analysis to estimate the sample size. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. The expected delta was 0.5 because [1) previous results published in ...; 2) of the following substantive reasons: ...]. Based on these parameters, a minimal sample size of 51 per group was estimated in order to reach 0.8 TPR for our design.

##### 1.2.1.2. Power curve {.unnumbered}

###### Study context {.unnumbered}

@@

###### Description {.unnumbered}

@@

###### Parameters {.unnumbered}

@@

###### How to report your sample size estimation {.unnumbered}

We used a power analysis to estimate the sample size. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. Because [1) we have no clear expectation of the magnitude of delta 2) we expected the delta to be around...], we include power calculations for delta ranging from 0.1 to 0.9. Based on these parameters, minimal sample sizes per group for different hypothetical effect sizes to reach 0.8 TPR can be found in Figure \@ref(fig:powercurve).

@@

#### 1.2.2. Bayesian {.unnumbered}

##### 1.2.2.1. Predetermined sample size with Bayes factor {.unnumbered}

###### Study context {.unnumbered}

@@

###### Description {.unnumbered}

The present method calculates the corresponding default Bayes factor for a *t*-test statistic with Cauchy prior distribution centered on zero with scale parameter 1/$\sqrt{2}$ for several sample sizes [the so-called Jeffrey-Zellner-Siow Bayes factor, see e.g., @rouder_bayesian_2009]. The function returns the optimal sample size needed to reach the TPR for a given Bayes factor threshold to detect an expected population effect size. If a range of possible population effect sizes are plausible under the given hypothesis, the function can calculate the optimal sample sizes for the given range of effect sizes and present the results in a figure (analogous to the Power Curve method).

###### Parameters {.unnumbered}

**Delta:** The expected population effect size or a range of expected effect sizes.\
**TPR:** The long-run probability of obtaining a Bayes factor at least as high as the critical threshold favoring superiority, given Delta.\
**Maximum N:** The maximum number of participants per group (both groups are assumed to have equal sample size).\
**Threshold:** Critical threshold for the Bayes factor. Three threshold levels are available in the app: 3, 6, and 10.\

###### How to use the package {.unnumbered}

```{r, eval = FALSE, tidy = TRUE, echo = TRUE}
SampleSizePlanner::ssp_bf_predetermined(
  tpr = 0.8, delta = 0.5, thresh = 10, max_n = 5000
  )
```

###### How to report your sample size estimation {.unnumbered}

We used the Jeffrey-Zellner-Siow Bayes factor method to estimate the sample size. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. The expected delta was 0.5 because [1) previous results published in ...; of the following substantive reasons: ...]. Our evidence threshold was 10. Based on these parameters, a minimal sample size of 105 per group was estimated in order to reach a 0.8 TPR for our design.

##### 1.2.2.2. Bayes Factor Design Analysis (BFDA) {.unnumbered}

###### Study context {.unnumbered}

@@

###### Description {.unnumbered}

The description of the BFDA method is functionally identical to the one provided in section 'Predetermined sample size with Bayes factor', but gains in TPR due to the addition of sequential testing. In the app, H~0~ and H~a~ indicate the proportion of times sequential testing leads to Bayes factors providing evidence with the given threshold for the null hypothesis and for the alternative hypothesis, respectively. For further reading, see @schonbrodt_bayes_2018 and @schonbrodt_sequential_2017.

###### Parameters {.unnumbered}

**Delta:** The expected population effect size.\
**TPR:** The long run probability of obtaining a Bayes factor at least as high as the critical threshold favoring superiority, given Delta.\
**Threshold:** Critical threshold for the Bayes factor. Three threshold levels are available in the app: 3, 6, and 10.\

###### How to use the package {.unnumbered}

```{r, eval = FALSE, tidy = TRUE, echo = TRUE}
SampleSizePlanner::ssp_bfda(
  tpr = 0.8, delta = 0.5, thresh = 10, n_rep = 1000
  )
```

###### How to report your sample size estimation {.unnumbered}

We used the BFDA method to estimate the sample size. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. The expected delta was 0.5 because [1) previous results published in ...; 2) of the following substantive reasons: ...]. Our evidence threshold was 10. Based on these parameters, a minimal sample size of 81 per group was estimated in order to reach a 0.8 TPR for our design.

## 2. Estimation {.unnumbered}

### 2.1. Frequentist {.unnumbered}

#### 2.1.1. Accuracy In Parameter Estimation (AIPE) {.unnumbered}

##### Study context {.unnumbered}

@@

##### Description {.unnumbered}

@@

#### Parameters {.unnumbered}

@@

##### How to report your sample size estimation {.unnumbered}

In order to estimate the sample size, we used the accuracy in parameter estimation [AIPE; @kelley_sample_2006] method. We aimed for a 95% confidence level, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. The desired width was 0.4 because [1) previous studies on this topic reported a similar region of practical equivalence; 2) of the following substantive reasons: ...]. We expected an underlying population effect size of 0.3, because [1) previous results published in ...; 2) of the following substantive reasons: ...]. Based on these parameters, a minimal sample size of 195 per group was estimated for our design.

#### 2.1.2. A Priori Precision (APP) {.unnumbered}

###### Study context {.unnumbered}

@@

##### Parameters {.unnumbered}

@@

###### How to report your sample size estimation {.unnumbered}

@@

### 2.2. Bayesian testimation {.unnumbered}

#### 2.2.1. Region of Practical Equivalence (ROPE) {.unnumbered}

###### Study context {.unnumbered}

@@

###### Parameters {.unnumbered}

**Delta:** The expected population effect size.\
**TPR:** The desired long run probability of having the HDI fully contained within the ROPE interval, given Delta.\
**EqBand:** The chosen ROPE interval.\

###### How to use the package {.unnumbered}

```{r, eval = FALSE, tidy = TRUE, echo = TRUE}
SampleSizePlanner::ssp_bf_predetermined(
  tpr = 0.8, delta = 0.5, thresh = 10, max_n = 5000
  )
```

###### How to report your sample size estimation {.unnumbered}

In order to estimate the sample size, we used the Region of Practical Equivalence [@kruschke_rejecting_2018] method. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. We consider all effect sizes below 0.2 equivalent to zero, because [1) previous studies reported a similar region of practical equivalence; 2) of the following substantive reasons: ...]. The expected delta was 0 because [1) we expected no difference between the groups]. Based on these parameters, a minimal sample size of 517 per group was estimated in order to reach a 0.8 TPR for our design.

# Summary {.unnumbered}

@@

# Authors contribution {.unnumbered}

@@

# Notes {.unnumbered}

## Glossary {.unnumbered}

*Accuracy in Parameter Estimation (AIPE):* A sample size estimation method used for parameter estimation. The approach aims to find the required sample size, such that the confidence interval has a certain expected width.\
*Priori Procedure (APP):* The approach aims to plan a sample size based on how close the researcher wishes both sample means to be to their respective population parameter, and how confident the researcher wants to be in this.\
*Bayesian inference:* A general framework for updating one's prior beliefs in light of new data.\
*Bayes Factor Design Analysis (BFDA):* This technique provides an expected sample size such that compelling evidence in the form of a Bayes factor can be collected for a given effect size with a certain long-run probability when allowing for sequential testing.\
*Testing vs. Estimation:* Two schools of inference, focusing on establishing whether or not an effect exists versus establishing the magnitude of an effect, respectively.\
*Equivalence band (EqBand):* The region of effect sizes considered practically equivalent to zero. In our paper, SESOI and ROPE are subsumed under EqBand.\
*Frequentist inference:* A general framework in which probabilities are defined as frequencies in hypothetical repeated events. In context of statistical testing, frequentist inference is concerned with long-run error rates of rejecting the null hypothesis for the observed or more extreme parameters in a given design when the model assumptions (e.g., independence of observations) are true.\
*Statistical power:* The long-run probability of finding a significant effect given a certain population effect size.\
*True positive rate (TPR):* The long-run probability of finding evidence for an effect, given that it exists. In our paper, statistical power is subsumed under TPR.\
*Classical power analysis:* This method is used to estimate the minimum sample size that a design needs to reach a certain level of statistical power, given a desired significance level and expected effect size.\
*Power-curve:* This curve shows how changes in effect size modify the statistical power of a test.\
*Region Of Practical Equivalence (ROPE):* The region of effect sizes considered practically equivalent to zero under the HDI-ROPE method.\
*Smallest Effect Size Of Interest (SESOI):* The region of effect sizes considered practically equivalent to zero under the TOST method.\
*Sequential testing:* The practice of incrementally testing as data comes in, typically until some pre-determined level of evidence is obtained.\
*Two One‐Sided Tests (TOST):* A frequentist statistical testing approach aimed at establishing equivalence between two groups.\
*Equivalence interval BF:* A Bayesian statistical testing approach aimed at establishing equivalence between two groups.\

@@

# References {.unnumbered}

>>>>

---
title             : "SampleSizePlanner: A Tool to Estimate and Justify Sample Size for Two-Group Studies"
shorttitle        : "SAMPLESIZEPLANNER"

@@

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
linkcolor         : "blue"
mask              : no
draft             : no
figsintext        : yes
numbersections    : false

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

# Introduction

@@

In the statistical literature, a few methods have been proposed to determine and justify sample size. In practice, however, authors are short of practical guides on how to navigate among the different sample size methods. The aim of our tutorial is to point out for each method the essential decision points that a researcher has to face during this process. We provide a short description of each method and the corresponding parameters, but we avoid listing their advantages and disadvantages. As there are disagreements between the experts of the field regarding the correct use of some of the methods, we intentionally try to remain impartial and do not favor any of the presented methods. Researchers who want to know more about each method can find a number of useful references in the description of the methods. We also provide a collection of ready-to-use analysis code and a ShinyApp that helps researchers use and report the main sample size estimation techniques for different scenarios. The tutorial is focused exclusively on the scenario of the comparison of two independent groups (i.e., the independent *t*-test design) with a one-sided test.

# Sample Size Determination and Justification

@@

## Choosing a method in case of sample size justification

@@

Within the realm of statistical testing, there are some other factors that affect the preferred inference tool: Do you prefer to test for equivalence (no difference in mean) or for superiority (mean of one group larger than mean of other group), are you interested in calculating a required sample size for a specific hypothetical effect size or for a range of possible values, and do you wish to employ sequential testing (applicable to Bayesian testing)? In case of testing, some of the methods are designed to find support for the null hypothesis (e.g., TOST, ROPE), while others are designed to find support for the alternative hypothesis (e.g., traditional null hypothesis testing), and some methods are designed to find support for either (e.g., BFDA). For frequentist estimation, the preferred inference tool might differ depending on whether we evaluate uncertainty for each group separately or jointly. We will describe these specific factors when we go into detail about each of the preferred methods. A flow-chart representing all of these choices is given in Figure \@ref(fig:flowchart).

@@

## How to use this guide

@@

For each method, only the main parameters can be adjusted with a certain range of values in the ShinyApp by using a slider. These parameters are presented in the text in bold. Other parameters are set to preset values in the application but can be adjusted in the accompanying R package to any sensible value. These parameters are highlighted in italics in the tutorial. Both the app and the package allow the users to save or copy a text template with the results of the sample size determination. We offer a list of possible justifications at the decision points for each method (indicated between square brackets), but users are able to provide their own justification as free-text. It is important to note that the listed justifications are meant to provide guidance for the user, and they are not sufficient without further details provided by the researcher in the context of the given study. For example, previously reported values should always be accompanied by a theoretical justification of why these values make sense. The provided justification text could serve as a stub for the description of the chosen sample size in a paper, a preregistration or registered report, or a grant proposal.

Throughout, we will use the example story of Mary the educational psychologist. Mary has come up with a new set of games that challenge spatial insight. She believes that distributed and targeted engagement with these games for a period of six months for children in the age range of 8 to 12 will lead to lasting improvements on their IQ score as measured through Raven's progressive matrices test (population mean 100, population SD 15). Mary collects data for a control sample that gets regular education and for an experimental sample and plans to compare those samples. For illustrative purposes, we will have Mary's hypothetical study goal depend on the earlier presented decisions to illustrate different scenarios for each sample size planning method. That means that in some of the upcoming examples, Mary expects a null effect and in others she expects a positive effect. We will also present a justification text based on Mary’s choices for each method.

@@

## 1. Testing

### 1.1. Effect size = 0

#### 1.1.1. Two One‐Sided Tests (TOST)

##### Study context

@@

##### Description

TOST is a frequentist equivalence testing approach that adopts two one-sided hypotheses to designate an interval hypothesis [@schuirmann_comparison_1987]. The lower and upper boundaries of the interval are determined by the equivalence band (i.e. SESOI) around the expected population effect size (e.g., 0). @lakens_equivalence_2018 lists several methods that can be used to determine the SESOI . In case of TOST, the two null hypotheses state that the effect size is equal to the lower and upper equivalence band values, whereas the alternative hypotheses state that the effect size is significantly smaller than the upper equivalence band value and significantly larger than the lower equivalence band value. In case both one-sided tests reject the null-hypothesis at a given significance level, the group means are considered to be practically equivalent. See @lakens_equivalence_2018, for further reading.

##### Parameters

@@

##### How to use the package

@@

##### How to report your sample size estimation

In order to calculate an appropriate sample size for testing whether the two groups are practically equivalent, we used the Two One-Sided Tests of Equivalence [TOST; @schuirmann_comparison_1987] method. We used an alpha of 0.05. We set the aimed TPR to be 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. We consider all effect sizes below 0.2 equivalent to zero, because [1) previous studies reported the choice of a similar equivalence band; 2) of the following substantive reasons: ...]. The expected delta was 0 because [1) we expected no difference between the groups]. Based on these parameters, a sample size of 429 per group was estimated in order to reach a TPR of 0.8 with our design.

#### 1.1.2. Equivalence interval Bayes factor

##### Study context

@@

##### Description

Equivalence interval Bayes factors contrast an equivalence hypothesis to a non-equivalence hypothesis and quantify the evidence with Bayes factors. Typically, H~0~ constitutes the equivalence interval (comparable to SESOI in the TOST framework), and H~a~ constitutes the complementary non-equivalence regions. Formally, the Bayes factor is calculated by dividing the fraction *posterior area inside the interval/posterior area outside the interval* (i.e., the posterior odds) by the fraction *prior area inside the interval/prior area outside the interval* (i.e., the prior odds). The resulting value quantifies how much more likely it is that the data occurred under a population effect size deemed 'equivalent' relative to the data having occurred under a population effect size deemed non-equivalent. The current implementation uses a default Cauchy prior on effect size with the possible scale parameters of medium (r = 1/$\sqrt{2}$), wide (r = 1), or ultra-wide (r = $\sqrt{2}$). For further reading, see @morey_bayes_2011, @van_ravenzwaaij_bayes_2019, and @linde_tendeiro_selker_wagenmakers_van_ravenzwaaij_2020.

##### Parameters

**Delta:** The expected population effect size.\
**TPR:** The desired long-run probability of obtaining a Bayes factor at least as high as the Threshold, given Delta.\
**EqBand:** The chosen width of the equivalence region.\
**PriorScale:** The scale of the Cauchy prior distribution. The PriorScale in the application can be set to: 1/$\sqrt{2}$, 1, and $\sqrt{2}$.\
*Threshold:* Critical threshold for the Bayes factor. The threshold level in the application can be set to 10, 6, or 3.\

##### How to use the package

```{r, eval = FALSE, tidy = TRUE, echo = TRUE}
SampleSizePlanner::ssp_eq_bf(
  tpr = 0.8, delta = 0,
  eq_band = 0.2, thresh = 10, prior_scale = 1/sqrt(2)
  )
```

##### How to report your sample size estimation

In order to estimate the sample size, we used the interval equivalent Bayes factor [@morey_bayes_2011; @van_ravenzwaaij_bayes_2019] method. We used a Cauchy prior distribution centered on zero with a scale of 1/$\sqrt{2}$. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. We consider all effect sizes below 0.2 equivalent to zero, because [1) previous studies reported the choice of a similar equivalence region; 2) of the following substantive reasons: ...]. The expected delta was 0 because [1) we expected no difference between the groups]. Our Bayes factor threshold for concluding equivalence was 10. Based on these parameters, a minimal sample size of 144 per group was estimated in order to reach 0.8 TPR for our design.

### 1.2. Effect size \>0

#### 1.2.1. Frequentist

##### 1.2.1.1. Classical power analysis

###### Study context

@@

###### Description

The classical power analysis approach allows one to calculate the required sample size in order to obtain a significant result for the null hypothesis test a certain proportion of times in the long run given an assumed population effect size.

###### Parameters

@@

###### How to use the package

@@

###### How to report your sample size estimation

We used a power analysis to estimate the sample size. We used an alpha of 0.05. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. The expected delta was 0.5 because [1) previous results published in ...; 2) of the following substantive reasons: ...]. Based on these parameters, a minimal sample size of 51 per group was estimated in order to reach 0.8 TPR for our design.

##### 1.2.1.2. Power curve

###### Study context

@@

###### Description

@@

###### Parameters

@@

###### How to report your sample size estimation

We used a power analysis to estimate the sample size. We used an alpha of 0.05. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. Because [1) we have no clear expectation of the magnitude of delta 2) we expected the delta to be around...], we include power calculations for delta ranging from 0.1 to 0.9. Based on these parameters, minimal sample sizes per group for different hypothetical effect sizes to reach 0.8 TPR can be found in Figure \@ref(fig:powercurve).

@@

#### 1.2.2. Bayesian

##### 1.2.2.1. Predetermined sample size with Bayes factor

###### Study context

@@

###### Description

The present method calculates the corresponding default Bayes factor for a *t*-test statistic with Cauchy prior distribution centered on zero with scale parameter of either 1/$\sqrt{2}$, 1, or $\sqrt{2}$ for several sample sizes [the so-called Jeffrey-Zellner-Siow Bayes factor, see e.g., @rouder_bayesian_2009]. The function returns the optimal sample size needed to reach the TPR for a given Bayes factor threshold to detect an expected population effect size. If a range of possible population effect sizes are plausible under the given hypothesis, the function can calculate the optimal sample sizes for the given range of effect sizes and present the results in a figure (analogous to the Power Curve method). This method is designed to determine the sample sizes for the existence of an effect (i.e., Delta > 0).

###### Parameters

**Delta:** The expected population effect size or a range of expected effect sizes.\
**TPR:** The long-run probability of obtaining a Bayes factor at least as high as the critical threshold favoring superiority, given Delta.\
**Maximum N:** The maximum number of participants per group (both groups are assumed to have equal sample size).\
**PriorScale:** The scale of the Cauchy prior distribution. The PriorScale in the application can be set to: 1/$\sqrt{2}$, 1, and $\sqrt{2}$.\
**Threshold:** Critical threshold for the Bayes factor. Three threshold levels are available in the app: 3, 6, and 10.\

###### How to use the package

```{r, eval = FALSE, tidy = TRUE, echo = TRUE}
SampleSizePlanner::ssp_bf_predetermined(
  tpr = 0.8, delta = 0.5, thresh = 10, max_n = 5000, prior_scale = 1/sqrt(2)
  )
```

###### How to report your sample size estimation

We used the Jeffrey-Zellner-Siow Bayes factor method to estimate the sample size. We used a Cauchy prior distribution centered on zero with a scale of 1/$\sqrt{2}$. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. The expected delta was 0.5 because [1) previous results published in ...; of the following substantive reasons: ...]. Our evidence threshold was 10. Based on these parameters, a minimal sample size of 105 per group was estimated in order to reach a 0.8 TPR for our design.

##### 1.2.2.2. Bayes Factor Design Analysis (BFDA)

###### Study context

@@

###### Description

The description of the BFDA method is functionally identical to the one provided in section 'Predetermined sample size with Bayes factor', but gains in TPR due to the addition of sequential testing. In the app, H~0~ and H~a~ indicate the proportion of times sequential testing leads to Bayes factors providing evidence with the given threshold for the null hypothesis and for the alternative hypothesis, respectively. Users of the Shiny app and R package should set Delta to 0 if they wish to determine the sufficient sample size for rejecting an effect, and use Delta > 0 if they wish to find support for the existence of an effect. For further reading, see @schonbrodt_bayes_2018 and @schonbrodt_sequential_2017.

###### Parameters

**Delta:** The expected population effect size.\
**TPR:** The long run probability of obtaining a Bayes factor at least as high as the critical threshold favoring superiority, given Delta.\
**PriorScale:** The scale of the Cauchy prior distribution. The PriorScale in the application can be set to: 1/$\sqrt{2}$, 1, and $\sqrt{2}$.\
**Threshold:** Critical threshold for the Bayes factor. Three threshold levels are available in the app: 3, 6, and 10.\

###### How to use the package

```{r, eval = FALSE, tidy = TRUE, echo = TRUE}
SampleSizePlanner::ssp_bfda(
  tpr = 0.8, delta = 0.5, thresh = 10, n_rep = 10000, prior_scale = 1/sqrt(2)
  )
```

###### How to report your sample size estimation

We used the BFDA method to estimate the sample size. We used a Cauchy prior distribution centered on zero with a scale of 1/$\sqrt{2}$. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. The expected delta was 0.5 because [1) previous results published in ...; 2) of the following substantive reasons: ...]. Our evidence threshold was 10. Based on these parameters, a minimal sample size of 81 per group was estimated in order to reach a 0.8 TPR for our design.

## 2. Estimation

### 2.1. Frequentist

#### 2.1.1. Accuracy In Parameter Estimation (AIPE)

##### Study context

@@

##### Description

@@

#### Parameters

@@

##### How to report your sample size estimation

In order to estimate the sample size, we used the accuracy in parameter estimation [AIPE; @kelley_sample_2006] method. We aimed for a 95% confidence level, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. The desired width was 0.4 because [1) previous studies reported the choice of a similar region of practical equivalence; 2) of the following substantive reasons: ...]. We expected an underlying population effect size of 0.3, because [1) previous results published in ...; 2) of the following substantive reasons: ...]. Based on these parameters, a minimal sample size of 195 per group was estimated for our design.

#### 2.1.2. A Priori Precision (APP)

###### Study context

@@

##### Parameters

@@

###### How to report your sample size estimation

@@

### 2.2. Bayesian testimation

#### 2.2.1. Region of Practical Equivalence (ROPE)

###### Study context

@@

###### Parameters

**Delta:** The expected population effect size.\
**TPR:** The desired long run probability of having the HDI fully contained within the ROPE interval, given Delta.\
**EqBand:** The chosen ROPE interval.\
**PriorScale:** The scale of the Cauchy prior distribution. The PriorScale in the application can be set to: 1/$\sqrt{2}$, 1, and $\sqrt{2}$.\

###### How to use the package

```{r, eval = FALSE, tidy = TRUE, echo = TRUE}
SampleSizePlanner::ssp_rope(
  tpr = 0.8, delta = 0.5, eq_band = 0.2, prior_scale = 1/sqrt(2)
  )
```

###### How to report your sample size estimation

In order to estimate the sample size, we used the Region of Practical Equivalence [@kruschke_rejecting_2018] method. We used a Cauchy prior distribution centered on zero with a scale of 1/$\sqrt{2}$. We set the aimed TPR at 0.8, because [1) it is the common standard in the field; 2) it is the journal publishing requirement]. We consider all effect sizes below 0.2 equivalent to zero, because [1) previous studies reported the choice of a similar region of practical equivalence; 2) of the following substantive reasons: ...]. The expected delta was 0 because [1) we expected no difference between the groups]. Based on these parameters, a minimal sample size of 517 per group was estimated in order to reach a 0.8 TPR for our design.

# Summary

@@

# Authors contribution

@@

# Notes

## Glossary

*Accuracy in Parameter Estimation (AIPE):* A sample size estimation method used for parameter estimation. The approach aims to find the required sample size, such that the confidence interval has a certain expected width.\
*Priori Procedure (APP):* The approach aims to plan a sample size based on how close the researcher wishes both sample means to be to their respective population parameter, and how confident the researcher wants to be in this.\
*Bayesian inference:* A general framework for updating one's prior beliefs in light of new data.\
*Bayes Factor Design Analysis (BFDA):* This technique provides an expected sample size such that compelling evidence in the form of a Bayes factor can be collected for a given effect size with a certain long-run probability when allowing for sequential testing.\
*Testing vs. Estimation:* Two schools of inference, focusing on establishing whether or not an effect exists versus establishing the magnitude of an effect, respectively.\
*Equivalence band (EqBand):* The region of effect sizes considered practically equivalent to zero. In our paper, SESOI and ROPE are subsumed under EqBand.\
*Frequentist inference:* A general framework in which probabilities are defined as frequencies in hypothetical repeated events. In the context of statistical testing, frequentist inference is concerned with long-run error rates of rejecting the null hypothesis for the observed or more extreme parameters in a given design when the model assumptions (e.g., independence of observations) are true.\
*Statistical power:* The long-run probability of finding a significant effect given a certain population effect size.\
*True positive rate (TPR):* The long-run probability of finding evidence for an effect, given that it exists. In our paper, statistical power is subsumed under TPR.\
*Classical power analysis:* This method is used to estimate the minimum sample size that a design needs to reach a certain level of statistical power, given a desired significance level and expected effect size.\
*Power-curve:* This curve shows how changes in effect size modify the statistical power of a test.\
*Region Of Practical Equivalence (ROPE):* The region of effect sizes considered practically equivalent to zero under the HDI-ROPE method.\
*Smallest Effect Size Of Interest (SESOI):* The region of effect sizes considered practically equivalent to zero under the TOST method.\
*Sequential testing:* The practice of incrementally testing as data comes in, typically until some pre-determined level of evidence is obtained.\
*Two One‐Sided Tests (TOST):* A frequentist statistical testing approach aimed at establishing equivalence between two groups.\
*Equivalence interval BF:* A Bayesian statistical testing approach aimed at establishing equivalence between two groups.\

@@

# References
